A Database Management System for a Geospatially-Enabled Interactive Game

### 2018-12-11 paper/MSDS7330DBMSforGeospatiallyEnabledGame.tex
Changed Appendix Placement
Final minor fixes

\section{Introduction}
Location based apps and services are becoming more and more popular. The game Pokémon Go, in particular, has gained over 44 million users and additionally incorporates augmented reality. We developed a database that can collect, store and aid in processing information for a game where players are collecting items in different tourist destinations to earn points. We called our game StoryPods. The game consist of players moving around the real world and collecting objects that gradually reveal parts of a story. The data is collected from a user interface operating on their cellphones. The emphasis of this work is to design a system that can efficiently collect and store geospatially enabled data, and use a database structure to manage the game rules and possible interactions among players.

@@

The story occurs in a science fiction setting that places the player in the role of an explorer. The player finds himself exploring ruins on another world that once housed an advanced civilization and is charged with learning how its members lived and what caused their disappearance. The player must then find objects, divided into three categories, that will divulge the necessary information about how this lost civilization lived and why it disappeared. Categories consist of Culture, Politics, and Technology and correspond to the nature of the player’s discoveries and tell the story of the civilization in tandem. For example, culture objects reveal how the civilization lived and what cultural tensions were present before they disappeared. Furthermore, the story arc of the civilization’s culture is told independently of the technology and politics story arcs, but with the common overarching plot of the civilization’s demise. An illustration of the story arc concept is presented below in \autoref{Tandem}.

@@

\begin{enumerate}
	\item Database Requirements for Working with Geospatial Data: It is necessary to understand data types, how they are stored, and how they may be obtained both from the player and by the player. 
	\item DBMS Options: It is necessary to survey what database management systems and existing software packages are available that work well with GPS data and are also capable of storing the required tables and enforcing game rules.
	\item Database Schema: It is imperative to know what entities and attributes are must be stored and what relationships need to be enforced in order to develop the conceptual design for the database.
	\item Client Side: In order to develop and test a prototype of the game, it is necessary to develop some kind of interface that interacts with the user and allows the user to play the game.
	\item Back End Services: In addition to the client-side application, it is necessary develop some application that implements the player’s actions and interacts with the database accordingly.

@@

\item The database must be able to store the player’s information, including login ID, email, and country of location.
\item The database must be able to represent GPS location data. An important aspect when dealing with geospatial data is how to represent the data efficiently in order to reduce network traffic and storage requirements. A geospatial location is defined by its latitude and longitude coordinates. There are different formats in which these coordinates can be represented. The most common format for many applications (such as Google Maps and Geographical Information Systems or GIS) is the signed degrees format in which latitudes range from -90 to 90, and longitudes from -180 to 180.
\item The database can update the player's location. For the purpose of supporting the gameplay, it is important to update player’s locations when they move to a new destination where collectible objects are available. The database system must ensure the consistency and accuracy of the data even when the information of many closely-located players is been updated frequently. 
\item The location data must be stored with enough precision to support the actions of collecting objects. Latitude and longitude data are more precise when they use more decimals. Using 8 decimal places would give an exact location within 1 millimeter. We can achieve our desired precision using 6 decimals. This information is important because it determines, along with the number of players, how much data is stored in the database. If the amount is very large, some sort of data compression would be necessary.
\item The system must have a function to calculate scores for each player, adding the value of the collected objects. The system also records subtotals for each object category and recognizes when thresholds are reached, triggering new story segments.
\item The database must be able to interface with the AWS platform where the data is stored, and the middle layer API.

@@

\item \textbf{PostgreSQL}. PostGIS is a spatial database extention for PostgreSQL. It adds support for geographic objects allowing location queries to be run in SQL. The three features that PostGIS delivers to PostgreSQL DBMS are spatial types, indexes and functions such as distance, area, union, intersection, and specialty geometry. AWS offers a purpose-built PostGreSQL database. Another advantage is that PostgreSQL is the recommended relational database for working with Python applications, which are a essential component of the middle layer.

@@

\begin{itemize}
\item	Once the data is uploaded, use simple SQL language to run complex queries related locations, objects, and other spatial data. 
\item Scale out the database in various ways: a. Manage a very large data size because it offers multi-TB capabilities, b. Handle a large amount of concurrency from the data, c. Use table-partitioning to break large data sets down into more manageable pieces.
\item Build custom functions using a language such as Python or R, which will be very useful to program the game requirements, rules and mechanisms into the database. 
\end{itemize}

@@

\begin{enumerate}
  \item Client-side application: We decided to support 2 major platforms that got widespread adoption: Android and iOS. For the proof-of-concept we focused on iOS. 
  \item Backend services, or middle layer: This component performs server-side operations, enforces gameplay rules and handles requests to our database. 
  \item Geospatially enabled database: Here all information regarding our players, objects and corresponding states is being kept. 
\end{enumerate}

@@

\begin{enumerate}[label=\Alph*]
\item REST over HTTP. This protocol is used to send messages from the client to the server sides and back
\item SQL. We use relational database and SQL over Application Layer to transform and query our data. 
\end{enumerate}

On the client side we have an iOS application that collects longitude and latitude of a smartphone and places objects around player. We calculate bearing, the horizontal angle between the direction of an object and player and distance between player and an object to place stories around given location.

@@

\subsubsection*{Players}
The UserData table contains login and personal information of all registered players taking part in the game. In the TempLocation table we store the player’s geographic location using the format provided by the PostGIS extension, which is a geographical object (latitude, longitude). We collect the player's location using the middle layer. We do keep history of a player’s location, but simply update the current location.

@@

We use Python to serve Application Program Interface (API) and PostGIS, a spatial database extender for PostgreSQL database for spatial queries.

\textbf{Player Registration:} Record user profile data (UserID, name, email, address and etc) from the client mobile device. The data stores in database at "userdata" table.  Input is in JSON format

@@

\textbf{Player location:} Instantaneously collect the player’s location data (latitude and longitude coordinates), and update this information in the database when the player moves according to the set thresholds (one update every 0.5 sec). The table has on record per user.

@@

\textbf{Collision report:} The database records new collected object for user and responds  to the client with new game score and revealing new story segments, which are triggered according to the segment threshold parameter in the database. Input and Response is in JSON format.

@@

\section{StoryPods Script}
\label{StoryPods Script}

>>>>

\section{Introduction}
Location based apps and services are becoming more and more popular. The game Pokémon Go, in particular, that additionally incorporates augmented reality has gained over millions of users. We developed a database that can collect, store and aid in processing information for a game where players are collecting items in different tourist destinations to earn points. We called our game StoryPods. The game consist of players moving around the real world and collecting objects that gradually reveal parts of a story. The data is collected from a user interface operating on their cellphones. The emphasis of this work is to design a system that can efficiently collect and store geospatially enabled data, and use a database structure to manage the game rules and possible interactions among players.

@@

The story occurs in a science fiction setting that places the player in the role of an explorer. The player finds himself exploring ruins on another world that once housed an advanced civilization and is charged with learning how its members lived and what caused their disappearance. The player must then find objects, divided into three categories, that will divulge the necessary information about how this lost civilization lived and why it disappeared. Categories consist of Culture, Politics, and Technology and correspond to the nature of the player’s discoveries and tell the story of the civilization in tandem. For example, culture objects reveal how the civilization lived and what cultural tensions were present before they disappeared. Furthermore, the story arc of the civilization’s culture is told independently of the technology and politics story arcs, but with the common overarching plot of the civilization’s demise. An illustration of the story arc concept is presented below in \autoref{Tandem}. The first episode of the story can be found in \autoref{StoryPods Script Episode 1} of the appendix.

@@

\begin{enumerate}
	\item Database requirements for working with geospatial data: It is necessary to understand data types, how they are stored, and how they may be obtained both from the player and by the player. 
	\item DBMS options: It is necessary to survey what database management systems and existing software packages are available that work well with GPS data and are also capable of storing the required tables and enforcing game rules.
	\item Database schema: It is imperative to know what entities and attributes must be stored and what relationships need to be enforced in order to develop the conceptual design for the database.
	\item Client side: In order to develop and test a prototype of the game, it is necessary to develop some kind of interface that interacts with the user and allows the user to play the game.
	\item Middle layer API: In addition to the client-side application, it is necessary develop some application that implements the player’s actions and interacts with the database accordingly.

@@

\item The database must be able to store the player’s information, including player's ID, email, and other attributes.
\item The database must be able to store GPS location data. An important aspect when dealing with geospatial data is how to represent the data efficiently in order to reduce network traffic and storage requirements. A geospatial location is defined by its latitude and longitude coordinates. There are different formats in which these coordinates can be represented. The most common format for many applications (such as Google Maps and Geographical Information Systems or GIS) is the signed degrees format in which latitudes range from -90 to 90, and longitudes from -180 to 180.
\item The database can update the player's location. For the purpose of supporting the gameplay, it is important to update player’s locations when they move to a new destination where collectible objects are available. The database system must ensure the consistency and accuracy of the data even when the information of many closely-located players is been updated frequently. 
\item The location data must be stored with enough precision to support the actions of collecting objects. Latitude and longitude data are more precise when they use more decimals. Using 8 decimal places would give an exact location within 1 millimeter. This information is important because it determines, along with the number of players, how much data is stored in the database. If the amount is very large, some sort of data compression would be necessary.
\item The system must have a function to calculate scores for each player, adding the value of the collected objects. The system also records subtotals for each object category and recognizes when thresholds are reached, triggering new story segments.
\item The database must be able to interface with the AWS platform where the data is stored, and the middle layer API.

@@

\item \textbf{PostgreSQL}. PostGIS is a spatial database extention for PostgreSQL. It adds support for geographic objects allowing location queries to be run in SQL. The three features that PostGIS delivers to PostgreSQL DBMS are spatial types, indexes and functions such as distance, area, union, intersection, and specialty geometry. AWS offers a purpose-built PostgreSQL database. Another advantage is that PostgreSQL is the recommended relational database for working with Python applications, which are a essential component of the middle layer.

@@

\begin{itemize}
\item	Once the data is uploaded, use simple SQL language to run complex queries related locations, objects, and other spatial data. 
\item Scale out the database in various ways: manage a very large data size because it offers multi-TB capabilities, handle a large amount of concurrency from the data, use table-partitioning to break large data sets down into more manageable pieces.
\item Build custom functions using a language such as Python or R, which will be very useful to program the game requirements, rules and mechanisms into the database. 
\end{itemize}

@@

\begin{enumerate}
  \item Client-side application: We decided to support 2 major platforms that got widespread adoption: Android and iOS. For the proof-of-concept we focused on iOS first. 
  \item Backend services, or middle layer: This component performs server-side operations, enforces gameplay rules and handles requests to our database. 
  \item Geospatially enabled database: Here all information regarding our players, objects and corresponding states is being kept. 
\end{enumerate}

@@

\begin{enumerate}[label=\Alph*]
\item REST over HTTP. This protocol is used to send messages from the client to the middle layer and back
\item SQL. We use relational database and SQL over Application Layer of the TCP/IP model to transform and query our data. 
\end{enumerate}

On the client side we have an iOS application that monitors longitude and latitude of a device and places objects around player. We calculate bearing, the horizontal angle between the direction of an object and player and distance between player and an object to place stories around given location.

@@

\subsubsection*{Players}
The UserData table contains ID and other attributes of all registered players taking part in the game. In the TempLocation table we store the player’s geographic location using the format provided by the PostGIS extension, which is a geographical point (latitude, longitude). We collect the player's location using the middle layer. We do not keep history of a player’s location.

@@

We use Python to serve Application Program Interface (API) and PostGIS, a spatial database extension of PostgreSQL for spatial queries. Endpoints of the middle perform following important functions:

\textbf{Player Registration:} Record user profile data (UserID, name, email, address and etc) of the client. The data stores in database at "userdata" table.  Input is in JSON format

@@

\textbf{Player location:} Instantaneously collect the player’s location data (latitude and longitude coordinates), and update this information in the database when the player moves according to the set thresholds (one update every 0.5 sec). The table has one record per user.

@@

\textbf{Collision report:} The middle layer stores a fact of a collision of a player with an object and responds with new player's score and reveals a new story segment. Input and Response is in JSON format.

@@

\section{StoryPods Script Episode 1}
\label{StoryPods Script Episode 1}
